%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentstyle[verbatimfiles,11pt,a4]{article}

\input{/ufs/gipe/src/ct/centaur/sources/ASDF/tolatex/ASF+SDF.tex}
\input{nodes}
\input{macros.tex}
\setlength{\textwidth}{15cm}
\setlength{\oddsidemargin}{0.46cm}
\setlength{\evensidemargin}{0.46cm}
\begin{document}
\bibliographystyle{alpha}
\title{\bf Type-checking revisited: Modular Error-handling}
\author{{\em T.B. Dinesh}, \\
        CWI, \\
        P.O. Box 4079, 1009 AB Amsterdam, The Netherlands\\
        e-mail: {\tt dinesh@cwi.nl}}
\maketitle
%
%\input{abstract}


\begin{abstract}

We study alternative ways in which the declarative
knowledge available in algebraic specifications of type-checkers
for a language, can be utilized in generated programming environments.
Algebraic specification formalisms are often based
on initial algebra semantics, in which it is easy to
introduce the so called {\em non-standard values}.
Using an {\em abstract-interpretation} style in such a
setting yields natural and concise type-checking specifications.
A separate module can then process the output
of such a type-checker to generate human-readable messages.

We use the type-checker specification of a toy language (Pico)
to illustrate the utility of the suggested style. 
We then discuss how this style facilitates {\em origin-tracking},
available in the \asdf\ system, to 
automatically identify the source of errors in a Pico program.

\end{abstract}

\input{pico-intro}
%\input{concl}	

\section{Concluding Remarks}

The style of type--checking in this document concentrates on specifying
only the necessary information, while still providing reasonably good
error messages. 
The style used makes use of the so called non-standard values of an
initial algebra specification to generate errors.
The result of type--checking is to effectively
form a conjunction of the abstract meanings of statements of the
language. All type correct statements evaluate to ``true'' while
an incorrect statement reduces to a structured error. 
This structured error can be used by a separate module
to generate human readble error messages.
Together with origin tracking, this provides 
information on the location of error in the source program automatically. 

%This is a typical example of how the specification
%style and the system can help process the declarative
%knowledge available in algebraic specification.

% This study also shows that the ASF+SDF meta-environment is
% a useful tool, which can be used to generate sophisticated programming
% environments and many necessary tools automatically.
% In particular, syntax-directed program construction, rapid prototyping,
% origin-tracking and effective source-level program analysis techniques
% such as visualization are available.

$$$$
{\bf Acknowledgements:} I would like to thank Arie van Deursen
for many discussions. Arie van Deursen and Susan Uskudarli
provided comments and constructive help with the drafts of the paper. 

%\bibliography{/ufs/gipe/lib/tex/bib,/ufs/gipe/spec/clax/doc/bib,bib}
\bibliography{bib,/ufs/gipe/lib/tex/bib}

\newpage
\appendix

\section{Pico Language}
\label{PICOSYN}

A Pico program consists of declarations followed by statements 
and is defined in Chapter~9 of \cite{BHK89}.
All variables are declared to be either of type natural or
of type string. Statements may be assignment statements,
if-statements and while-statements.
Expressions may be a single identifier, addition or subtraction
of natural numbers, or concatenation of strings.

The imported modules define the lexicals identifiers,
constants for natural numbers and strings used in the language.

The variables defined in this module are used in the
equations of modules that import Pico-syntax 
(e.g., Pico-typecheck modules).

\input{Pico-syntax}

\newpage

Example Pico program:

\verbatimfile{small.pico}

Specifying a type-checker for Pico was done using a table
for type-environments TENV, similar to list representation used below:

\input{Type-environments}

\end{document}
