%%
%%    Meta-Environment - An environment for language prototyping.
%%    Copyright (C) 2000  Stichting Mathematisch Centrum, Amsterdam, The Netherlands. 
%%
%%    This program is free software; you can redistribute it and/or modify
%%    it under the terms of the GNU General Public License as published by
%%    the Free Software Foundation; either version 2 of the License, or
%%    (at your option) any later version.
%%
%%    This program is distributed in the hope that it will be useful,
%%    but WITHOUT ANY WARRANTY; without even the implied warranty of
%%    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
%%    GNU General Public License for more details.
%%
%%    You should have received a copy of the GNU General Public License
%%    along with this program; if not, write to the Free Software
%%    Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307 USA
%%
\section{Tree Representation Formalism}
\label{treerepr}

In the current version of the \metaenv\ the abstract syntax trees are
represented by means of VTP (Virtual Tree Processing formalism)
offered by \centaur.  There are two problems connected to VTP: it is
hard to learn programming in VTP and VTP does not offer enough
facilities to prevent illegal access to constructed trees. This latter
drawback causes a number of the maintenance problems in the \metaenv.

These ``VTP-problems'' led to the development of an alternative
formalism to represent syntax trees: \asfix.  This \asfix\ formalism
is an instantiation of a generic {\em annotated term} format:
\aterms~\cite{BKOV97}.  

\aterms\ will be used to represent structured information to be
exchanged between a heterogeneous collection of tools.  The \aterms\
format should be independent of any specific tool, but it should be
capable of representing {\em all} data that is exchanged between
tools.  Consider the following example \aterms:
\begin{tabbing}
annotation\ \ \ \ \= f(123)\{\=xxxx\kill
constant\>abc\\
numeral\>123\\
literal\>"abc" "123"\\
list\>[] [1, "abc", 3] [1, 2, [3,4], 5]\\
functions\>f("abc")\\
annotation\>f(123)\{color("red"),path([0,2,1])\}
\end{tabbing}

\subsection{\asfix}

These \aterms\ are sufficient to encode parse trees (including
optional annotations) for programs or specifications in any language.
\asfix\ is an instance of \aterms\ used for representing the {\em
parse trees} of \asdf\ specifications.  Parse trees mean that all
keywords, whitespace, comments, etc.\ are preserved in the tree
representation.  A self descriptive representation is used, i.e., each
application of a syntax rule contains a copy of that rule (in the
implementation these are all shared).

\subsection{Implementation}

One of the strong points of \aterms\ are the powerful access functions
to retrieve information from the terms and the construction functions
to build terms. 
These functions are formally specified in \asdf~\cite{BKOV97} and
this formal specification is used to make a library of C functions
to manipulate terms.


