\newcommand{\hint}{\par\noindent\textbf{Hint:~}}

These exercises are meant to familiarize new users with the \ASmetaenv{}. They focus on the use of the \ASmetaenv{} as a language design tool. A \emph{Guided Tour} is also available as a part of the user-manual. The user-manual can be found at \url{http://www.cwi.nl/projects/MetaEnv/meta/doc/manual.ps.gz}, or in the \texttt{meta/doc} directory from the source distribution at \url{http://www.cwi.nl/projects/MetaEnv/asfsdf-meta}.

\section{Getting acquainted}

\begin{itemize}

\item Copy the Pico specification and start the \ASmetaenv{}:
\begin{verbatim}
 > cp -r /source/location/asfsdf-meta/meta/demo/pico .
 > cd pico
 > meta
\end{verbatim}
\noindent Where \texttt{/source/location} is the path to the source distribution of the \ASmetaenv{}. 
 
\item Open the Pico-Booleans specification (\texttt{File|Open}\footnote{This notation means: click on the \texttt{File} pull-down menu and choose the \texttt{Open} button.}) and have a look at the syntax
and equations by selecting the \texttt{Edit Syntax} and \texttt{Edit
Equations} options from the popup-menu. The popup-menu can be activated
by clicking on the Pico-Booleans rectangle in the graph-browser.

\item Create a new term by selecting the \texttt{New Term} option from
the Pico-Booleans popup-menu. Choose any name you like, e.g. \texttt{bool.trm}.

\item Enter the following term over the Pico-Booleans specification:
\begin{verbatim}
  true & false
\end{verbatim}
\hint The use of \emph{Search \& Replace} and \emph{Auto-Completion} in XEmacs is not supported by the \ASmetaenv{} and will cause unexpected behaviour.

\item Parse this term by selecting the \texttt{Parse} option from the
\texttt{Meta-Environment} menu (in the editor).

\item Click on \texttt{true} in the editor and observe the focus that is
now placed around the term. Also note that the sort (\texttt{PICO-BOOL})
is displayed in the bottom status area of the editor.

\item Reduce the term \texttt{true \& false} by selecting the
\texttt{Reduce} option from the menu. Inspect that the result is
indeed what you expected it to be, and close the reduct-window.

\item Change the subterm \texttt{false} into \texttt{true} and select
\texttt{Parse} and then \texttt{Reduce} again. 
\end{itemize}

\section{Extending the booleans}

\begin{itemize}

\item Extend the syntax of Pico-Booleans by adding a syntax rule for
the logical implication ($\Rightarrow$).

\item To activate the new syntax, re-parse it (\texttt{Meta-Environment|Parse}).

\item Inspect the rule you just entered, observing the focus and the
sorts displayed at the bottom of the editor.

\item Open the Pico-Booleans equations (if you haven't done so already).

\item Add the equations (semantic rules) for the $\Rightarrow$ operator.

\item To activate the new equations, \texttt{Parse} them.

\item Open a term-editor (graph-browser: \texttt{New Term}) over the
Pico-Booleans and test the $\Rightarrow$ semantics by entering and
reducing several terms.
\end{itemize}

\hint Close an editor after you have finished editing. This will save some memory, resulting in faster turnaround times.

\section{The Pico programming language}

\subsection{The Pico definition}
\begin{itemize}
\item Open the Pico-Syntax specification. Observe the modular design of the Pico language. 
\item Open the term \texttt{fac.trm}. This is a Pico program which computes the factorial.
\item Press \texttt{Parse} and click around in the term to get acquainted with the structure of a Pico program.
\item Also have a look at the Pico-Syntax syntax definition and some of the modules imported by it. 
\item Open the Pico-typecheck module. This module defines a typechecker for Pico programs. 
\item Test the typechecker by opening the term \texttt{factc.trm} and reducing it. Create an error in \texttt{factc.trm}, for example by changing the type of a variable from \texttt{natural} to \texttt{string} and \texttt{Reduce} again.
\item Open the Pico-eval module. This module defines an interpreter for Pico  programs.
\item Test this specification by reducing the term \texttt{facev.trm}. 
\end{itemize}

\subsection{Optional \texttt{else} branches}
\begin{itemize}
\item Adapt the syntax definition of Pico such that a conditional statement can  be either an \texttt{if-then-fi} or an \texttt{if-then-else-fi}. Note that there are several ways to do this. Choose one you like.
\item You have now broken the typechecker and the interpreter. After you have figured out why they are broken, fix them. 
\end{itemize}

\hint Remember to press \texttt{Parse} in order to effectuate your changes.
\hint If you make a mistake that causes the \ASmetaenv{} to loop continuously, quit (\texttt{File|Quit}) and restart it. If that does not work, type \verb#CTRL+\# in the terminal where you typed \texttt{meta}.

\subsection{A new basic type for Pico}
\begin{itemize}
\item Notice how Pico has only two basic types: \texttt{natural} and \texttt{string}. Add syntax for a new type: \texttt{boolean}.
\item Extend the Pico expression syntax (sort \texttt{EXP}) with Boolean constants (\texttt{PICO-BOOL-CON}) and four operations: \texttt{\&} , \texttt{|}, \texttt{==} and \texttt{!=}. Also express the relative priorities for the newly added operators. The \texttt{\&} and \texttt{|} operators should be \texttt{left} associative, while \texttt{==} and \texttt{!=} should be made \texttt{non-assoc}.
\item Extend the typechecker to cope with these kind of expressions. The operations \texttt{\&} and \texttt{|} only accept \texttt{boolean} operands. The \texttt{==} and \texttt{!=} operators should enforce that both operands are of \emph{equal} type. By doing this you are \emph{overloading} these two Pico operators for the three basic types. \hint Use the \texttt{compatible} function defined in \texttt{Types} to do this.
\item Extend the interpreter to cope with these kind of expressions. \hint First add \texttt{PICO-BOOL} as a new type of \texttt{VALUE} in the Values module. \hint Don't forget to look at the initialization of a variable that is done during evaluation of the declaration.
\end{itemize}

\subsection{Boolean conditional expressions}

The Pico typechecker currently enforces that the type of the conditional expression is \texttt{natural}. You have just added the Pico type \texttt{boolean}. Let's change the conditional to use booleans instead of naturals. 

\begin{itemize}
\item Change the typechecker such that type \texttt{boolean} is expected in the \texttt{if-then-fi} and the \texttt{if-then-else-fi}.
\item Change the evaluator such that the \texttt{then} branch is chosen when the expression evaluates to \texttt{true} and the (optional) \texttt{else} branch otherwise.
\end{itemize}
