\documentstyle[12pt,verbatimfiles]{article}

\raggedbottom
\setlength{\textwidth}{16cm}
\setlength{\textheight}{22cm}
\setlength{\oddsidemargin}{0.46cm}
\setlength{\evensidemargin}{0.46cm}

\newcommand{\asdf}{\mbox{ASF{\tt +}SDF}}


\newcommand{\actmod}[2]{
  \subsection{#2}
    \subsubsection{MetaNotation Module}
       \verbatimfile{#1/#2.asd}
    \subsubsection{Generated \asdf}
       \verbatimfile{#1/#2.syn}
       \verbatimfile{#1/#2.eqs}
    \newpage
}

\newcommand{\asdfmod}[2]{
  \subsubsection{#2}
  \verbatimfile{#1/#2.syn}
  \verbatimfile{#1/#2.eqs}
}

\newcommand{\asdfmodsyn}[2]{
  \subsubsection{#2}
  \verbatimfile{#1/#2.syn}
}

\begin{document}

\section*{Starting up the ASF+SDF tool for Action Semantics:}
\subsection*{Version 1.03; Draft!}


 To be written.

 You'll need  experience with ASF+SDF. Get the
 ASF+SDF user manual (ftp ``ftp.cwi.nl: /pub/gipe'').


 \begin{itemize}  
 \item{} 
   Start system in meta-notation directory; type ``ctasdf 120''
  \item
    Click ``ASF+SDF'' button.
    Now the Meta-Environment pops up, and it starts loading
    module MetaNotation, and the modules needed for it.
  \item
    Open a term editor over module MetaNotation,
    e.g. ``ActionNotation.asd''.
  \item
    Click somewhere, try to make some changes.
  \item
    To see the generated ASF+SDF modules for it, press ``Show All''
    button

    This will load the ASF+SDF modules needed to do the translation.
    The ASF+SDF meta-environment will now process all these modules
    for the first time; this takes a while. Please be patient.

  \item
    Before pushing any other button, first push Initialize.

  \item
    Button Dump SDF dumps the SDF part on file

    (try with ActionNotation.asd) The dumped file is in the same directory 
    as the original file; you can take a look at it using vi (emacs).

  \item
    Button Dump equations dumps the equations part on file

   (try with ActionNotation.asd)
  \item 
    Button Dump All dumps both

   (try with ActionNotation.asd)

  \item
    Button nested dump will automatically start term editors
    for {\em all\/} modules needed by this particular module.
    It will generate ASF+SDF for {\em all\/} these modules, and
    dump everyone on file.

    (try with ActionNotation.asd)

  \item
    Button Load will start a new ASF+SDF environment, in which
    the generated ASF+SDF is loaded.
    This will perform a considerable number of checks on the
    generated modules.
    
    (try with ``Pico-Semantic-Functions.asd'' as edited file).

  \item
    Button dump and load dumps and loads.

  \item
    The generated Meta-Environment is not only useful for checking;
    it can also be applied to {\em execute\/} semantic equations.
    E.g., open a term over ``Pico-Semantic-Funtions'', e.g.
    ``Pico-Semantic-Functions.trm''.
    Press reduce.
    You'll see the corresponding action term.

 \end{itemize}
  


\end{document}




